### 2020-06-17 report/sections/abstract.tex
Report: update AIC

\noindent
Multiparty Session Types (MPST)
is a typing discipline for concurrent
processes that communicate via
message-passing. MPST theory can ensure
communication safety properties and
protocol conformance.
Existing work in session-typed
web development over WebSocket transport
is incompatible with modern web
programming practices and are limited
to supporting communication protocols
that implement the server-centric network
topology.

@@

\codegen provides
developers with TypeScript APIs generated
from a communication protocol specification 
based on multiparty session type theory.
Our work is compatible with modern web programming
industrial practices. The generated APIs
build upon TypeScript concurrency primitives,
complement the event-driven style of programming
in full-stack web development,
and are compatible
with the Node.js runtime and the React.js framework
for server-side and browser-side endpoints respectively.
We evaluate the expressiveness of \codegen for modern
web programming
using case studies of protocols found in web-services,
and analyse the overhead of our implementation through 
running micro-benchmarks against a baseline implementation.

>>>>

\noindent
Modern web programming
involves coordinating interactions
between browser clients and a web server.
Typically, the interactions are
informally described, which makes
it difficult to verify communication
correctness in the distributed system.

Multiparty Session Types (MPST)
is a typing discipline for concurrent
processes that communicate via
message-passing. MPST theory can ensure
communication safety properties and
protocol conformance.
Existing work in session-typed
web development over WebSocket transport
is incompatible with modern web
programming practices and are limited
to supporting communication protocols
that implement the server-centric network
topology.

@@

\codegen provides
developers with TypeScript APIs generated
from a communication protocol specification 
based on multiparty session type theory.
Our work is compatible with modern web programming
industrial practices. The generated APIs
build upon TypeScript concurrency practices,
complement the event-driven style of programming
in full-stack web development,
and are compatible
with the Node.js runtime and the React.js framework
for server-side and browser-side endpoints respectively.
We evaluate the expressiveness of \codegen for modern
web programming
using case studies of protocols found in web-services,
and analyse the performance overhead
through 
running benchmarks against a baseline implementation.

### 2020-06-17 report/sections/evalconcl/concl/contributions.tex
Report: update AIC

We motivate our API generation approach from 
\cite{Hybrid2016,PureScript2019} to generate TypeScript APIs
from a Scribble protocol specification.
Developers write their full-stack applications
by implementing the generated APIs, and can enjoy 
communication safety guarantees by construction.
We explain how our session type encodings target the Node.js
server-side runtime and React.js framework in
\cref{chap:node,chap:react} for server-side
and browser-side endpoints respectively.
In particular, we present a novel approach for integrating
session types into web-based GUI programming based on
translating the theory on model types \cite{MVU2020}
to idiomatic practices on the React.js framework.
With respect to session type theory, implementations using
the generated APIs statically enjoy linear channel usage
guarantees and affine channel usage guarantees for back-end
and front-end targets respectively.

Compared to previous work 
\cite{Exceptional,PureScript2019,MVU2020,LINKS} on
session-typed web development, 
we are not only able to
\textbf{statically} provide the same level of communication safety
guarantees from multiparty session type theory, but we do so
using modern web programming practices to increase
the relevance and usability of our work in industry:
our work targets TypeScript, Node.js and React.js,
and we explain in \cref{chap:ext} how we support advanced
web development idioms (such as asynchronous implementations)
and handle premature disconnections gracefully.

@@

% PLACES
Our work has received positive feedback from academia.
The project has been accepted and published as 
\emph{Generating Interactive WebSocket Applications in TypeScript}
in Volume 314 of the Electronic Proceedings in Theoretical
Computer Science (EPTCS) \cite{PLACES2020}.
We were also invited to present this work
at the 12th International Workshop on
Programming Language Approaches to Concurrency- \& 
Communication-cEntric Software (PLACES 2020).

>>>>

We motivate our API generation approach from 
\cite{Hybrid2016,PureScript2019} to generate TypeScript APIs
from a Scribble protocol specification.
Developers write their full-stack applications
by implementing the generated APIs, and enjoy 
communication safety guarantees by construction.
APIs generated for server-side endpoints
are compatible with the Node.js runtime,
as detailed in \cref{chap:node}.
For browser-side endpoints,
we present a novel approach for integrating
session types into web-based GUI programming based on
translating the theory on model types \cite{MVU2020}
to idiomatic practices on the React.js framework;
this is highlighted in \cref{chap:react}.
With respect to session type theory, implementations using
the generated APIs statically enjoy linear channel usage
guarantees and affine channel usage guarantees for back-end
and front-end targets respectively.

Compared to previous work 
\cite{Exceptional,PureScript2019,MVU2020,LINKS} on
session-typed web development, 
we are not only able to
statically provide the same level of communication safety
guarantees from multiparty session type theory, but we do so
using modern web programming practices to increase
the relevance and usability of our work in industry:
our work targets TypeScript, Node.js and React.js,
and we explain in \cref{chap:ext} how we support advanced
web development idioms (such as asynchronous implementations)
and handle premature disconnections gracefully.

@@

% PLACES
Our work has received positive feedback from academia.
The project has been accepted and published as 
\emph{Generating Interactive WebSocket Applications in TypeScript}
in Volume 314 of the Electronic Proceedings in Theoretical
Computer Science (EPTCS) \cite{PLACES2020}.
We were also invited to present a part of this work
at the 12th International Workshop on
Programming Language Approaches to Concurrency- \& 
Communication-cEntric Software (PLACES 2020).

### 2020-06-17 report/sections/evalconcl/concl/future.tex
Report: update AIC

\paragraph{Explicit Connection Actions}
Hu and Yoshida \cite{ExplicitConnections} extends 
multiparty session type theory
with \textit{explicit connection actions} to support protocols 
with optional participants.
Supporting optional participants is a relevant concept
for web applications, particularly for browser-side endpoints.

@@

\paragraph{Transport Abstraction}
The MPST-based web development framework presented by 
King et al. \cite{PureScript2019} defines the session runtime
in a way that parameterises the transportation mechanism.
The developer can run the session over different transportation,
provided that the custom transport implements the behaviour
of the \texttt{Transport} type class.

>>>>

\paragraph{Explicit Connection Actions}
Hu and Yoshida \cite{ExplicitConnections} extended 
MPST theory
with \textit{explicit connection actions} to support protocols 
with optional participants.
Supporting optional participants is a relevant concept
for web applications, particularly for browser-side endpoints.

@@

\paragraph{Transport Abstraction}
The MPST-based web development framework presented by 
King et al. \cite{PureScript2019} parameterised
the transport mechanism in the session runtime definition.
The developer can run the session over different transportation,
provided that the custom transport implements the behaviour
of the \texttt{Transport} type class.

### 2020-06-17 report/sections/introduction/contributions.tex
Report: update AIC

We present an implementation of the MPST framework for developing
interactive web applications over WebSocket transport.
We also present a theory of routed multiparty session types,
which forms the basis for supporting arbitrary communication structures
over a server-centric WebSocket-based network topology.
Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} 
requires the 
developer to implement their application using tools that are 
unconventional for web development, 
and only support communication protocols that adhere to the
WebSocket network structure.
To the best of our knowledge, this is the first work on multiparty session
type-safe web development that provides the same communication safety 
guarantees using familiar industry tools in TypeScript and React,
and supports peer-to-peer browser interactions over WebSocket transport.

In \textbf{\cref{chap:background}}, 
we introduce the core session type theory
and provide an overview of related work on integrating session types into
application development. We also introduce TypeScript and highlight some 
interesting features and caveats in the language.

>>>>

We present an implementation of the MPST framework for developing
interactive web applications over WebSocket transport.
We also present a theory of routed multiparty session types,
which forms the basis for supporting arbitrary communication structures
over a server-centric WebSocket-based network topology.
Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} 
requires the 
developer to implement their application using tools that are 
unconventional for web development, 
and only support communication protocols that adhere to the
WebSocket network structure.
To the best of our knowledge, this is the first work on multiparty session
type-safe web development that provides the same communication safety 
guarantees using familiar industry tools in TypeScript and React,
and supports peer-to-peer browser interactions over WebSockets.

In \textbf{\cref{chap:background}}, 
we introduce the core session type theory
and provide an overview of related work on integrating session types into
application development. We also introduce TypeScript and highlight the 
defining features of the language.

### 2020-06-17 report/sections/introduction/motivation.tex
Report: new introduction
Report: update AIC

%\subsubsection{New Structure (similar to ECOOP2017):}
%\begin{itemize}
%
%\item \textbf{Section 1.1:}
%
%Interactive webapps over WebSockets, harder to guarantee communication safety
%
%\item
%Example protocol similar to travel agency  -- motivate that it is difficult to manually verify communication safety
%
%\begin{itemize}
%\item traveller enquires availability, flight server replies yes/no then traveller confirms/denies.
%\item The flight server will reserve the seat until the traveller denies.
%\item Source of hidden bugs -- traveller may enquire twice (e.g. click twice), but server processes one enquiry, so one seat is held up for no reason
%\end{itemize} 
%
%\item MPST framework can be applied for webapps.
%This is solved by existing work PureScript and MVU2020.
%
%\item \textbf{Problem1}: 
%
%existing work provides a workflow for session typed web development, but requires the developer to compromise by adapting to programming paradigms and practices not conventional for modern web programming. E.g. functional paradigm, sequential UI framework, PureScript and LINKS not interoperable with the ever-growing universe of JavaScript libraries. These compromises limit the usability of MPST framework
%for web development.
%
%\item \textbf{PRoblem2}: 
%
%existing work based on WebSocket transport, which
%is simple to reason about. But WebSocket transport enforces a server-centric
%network configuration, so existing work isn't expressive enough for
%more general protoocls. Suppose traveler contacts agency first to get a quote and then compare (similar to 2-buyer). Existing work cannot implement this.
%
%\item \textbf{Section 1.2: }
%
%Our objective is to present a novel variant of MPST framework
%that is compatible / supports the flexibility of modern web programming.
%We address the two problems in two parts:
%
%\item We solve problem 1 in part 1: 
%
%SessionTS is API generation workflow based on MPST, which
%provides developers with same communication safety guarantees in a way that is idiomatic for modern web programming. We use industry tools (TS/Node/REact), compatible with async logic, error handling specific to web (e.g. client disconnection)
%
%\item We solve problem 2 in part 2:
%
%develop novel theory for routed MPST which lets us implement non-server-centric commuication structures over a server-centric network topology, prove
%that our theory preserves properties like communication safety, deadlock freedom, etc.
%
%\item Section 1.3: diagram similar to FZ for outline
%
%\end{itemize}
%
%% Aim of introductio: motivate what is the problem 
%%(session types fix web, PureScript and LINKS have two problems: 
%%implementation -- not using mainstream stuff, 
%%theory -- websocket transport restricts protocols)
%

@@

\paragraph{The need for communication safety 
in interactive web applications}

@@

We can implement the flight booking service as a web
application.
The \trole{FlightServer} can be a web server
running on the \textit{Node.js} JavaScript runtime \cite{nodejs}.
The \trole{Traveller} can run on the web browser
as a \textit{single-page application} (SPA) written
in a popular framework like \textit{React.js} \cite{React}.
SPAs feature a single HTML page and dynamically renders
content via JavaScript in the browser.
The \trole{Traveller} connects to the \trole{FlightServer}
using a WebSocket connection and they enjoy
bi-directional communication to execute the flight
booking operations outlined above.

Whilst WebSockets make this web-based implementation possible, 
it introduces the developer to a new family of communication errors
(in addition to the usual testing for application logic correctness), 
even for this simple example:
we highlight just a few:

@@

\paragraph{Formalising Interactions with
Multiparty Session Types}
\textit{Multiparty Session Types} (MPST) \cite{MPST} 
provide a framework for formally specifying 
a structured communication pattern 
between concurrent processes and 
verifying implementations for 
correctness with respect to the communications aspect. 
By specifying the client-server interactions of 
our flight booking service as a protocol 
and verifying the implementations against 
the protocol for well-formedness, 
MPST theory guarantees well-formed 
implementations to be free from communication errors.

\paragraph{Limitations of State-of-the-Art Session-Typed Web Development}
Existing work \cite{MVU2020,PureScript2019} that 
adapt the MPST framework for web development
acknowledge the limitations of JavaScript 
-- the language of the browser --
in providing static type-level guarantees
for communication safety, 
and proceed to use
languages equipped with stronger type systems that 
compile to JavaScript instead.

@@

The current state-of-the-art succeeds in \textbf{statically}
providing communication safety guarantees,
but requires the developer to compromise by adapting
to programming paradigms and practices that are not
conventional for modern web programming.
Developers that use \cite{MVU2020} are required
to implement their web applications in a \textit{functional} 
paradigm using the \textit{Links} 
web programming language \cite{LINKS}.
\cite{PureScript2019} also endorses
functional web programming by requiring developers to
write PureScript \cite{PureScript} applications, but also
enforces user interfaces to be written using 
sequential UI frameworks.

PureScript and Links are not interoperable
with the ever-growing universe of JavaScript libraries for
web development.
It is also unclear whether these tools support idiomatic
practices in event-driven programming, such as callbacks
and asynchronous implementations.
Whilst communication safety guarantees provided by
\cite{PureScript2019,MVU2020} are based on
canonical MPST theory, they are not contextualised
for the dynamic web-based environment.
Going back to the flight booking service as an example,
\cite{PureScript2019,MVU2020} do not handle premature
disconnections; if the \trole{Traveller} receives a quote
and closes their browser prematurely and the \trole{FlightServer}
isn't notified of this event, the reserved seat will not be
released.
These compromises limit the usability of the MPST framework
in modern web programming.

>>>>

@@

\paragraph{Communication Safety 
in Interactive Web Applications}

@@

We can implement the flight booking service as a web
application.
\trole{FlightServer} can be a web server
running on the \textit{Node.js} JavaScript runtime \cite{nodejs}.
\trole{Traveller} can run on the web browser
as a \textit{single-page application} (SPA) written
in a popular framework like \textit{React.js} \cite{React}.
SPAs feature a single HTML page and dynamically renders
content via JavaScript in the browser.
The \trole{Traveller} connects to the \trole{FlightServer}
using a WebSocket connection and they enjoy
bi-directional communication to execute the flight
booking operations outlined above.
We visualise the interactions in \cref{fig:flightbook}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.7\textwidth]{FlightBookingService}
\captionof{figure}{Message-Passing Architecture of Web-Based 
Flight Booking Service}
\label{fig:flightbook}
\end{figure}

Whilst WebSockets make this web-based implementation possible, 
it introduces the developer to a new family of communication errors
(in addition to the usual testing for application logic correctness), 
even for this simple example:

@@

\paragraph{Formalising Interactions with
Multiparty Session Types}
\textit{Multiparty Session Types} (MPST) \cite{MPST} 
provide a framework for formally specifying 
a structured communication pattern 
between concurrent processes and 
verifying implementations for 
correctness with respect to the communications aspect.

By specifying the client-server interactions of 
our flight booking service as a protocol 
and verifying the implementations against 
the protocol for well-formedness, 
MPST theory guarantees well-formed 
implementations to be free from communication errors.

\paragraph{Limitations of State-of-the-Art Session-Typed Web Development}
Existing work \cite{MVU2020,PureScript2019} that 
adapt the MPST framework for web development
acknowledge the limitations of \textit{JavaScript} 
-- the language of the browser --
in providing static type-level guarantees
for communication safety, 
and proceed to use
languages equipped with stronger type systems that 
compile to JavaScript instead.

@@

The current state-of-the-art succeeds in statically
providing communication safety guarantees,
but requires the developer to compromise by adapting
to programming paradigms and practices that are not
conventional for modern web programming.
Developers that use \cite{MVU2020} are required
to implement their web applications in a \textit{functional} 
paradigm using the \textit{Links} 
web programming language \cite{LINKS}.
\cite{PureScript2019} also endorses
functional web programming by requiring developers to
write PureScript \cite{PureScript} applications, but also
enforces user interfaces to be written using 
sequential UI frameworks.

PureScript and Links are not interoperable
with the ever-growing universe of JavaScript libraries for
web development.
It is also unclear whether these tools support idiomatic
practices in event-driven programming, such as callbacks
and asynchronous implementations.
Whilst communication safety guarantees provided by
\cite{PureScript2019,MVU2020} are based on
canonical MPST theory, they are not contextualised
for the dynamic web-based environment.
Going back to the flight booking service as an example,
\cite{PureScript2019,MVU2020} do not handle premature
disconnections; if the \trole{Traveller} receives a quote
and closes their browser prematurely and the \trole{FlightServer}
is not notified of this event, 
the reserved seat will not be released.
These compromises limit the usability of the MPST framework
in modern web programming.

### 2020-06-17 report/sections/introduction/objectives.tex
Report: update AIC

The objective of this project is to provide
developers with a development workflow
that provides communication safety guarantees in modern
web programming through multiparty session types.

We tackle the limitations of the current
state-of-the-art with a two-step strategy:

@@

Developers specify their communication protocol using
the Scribble protocol description language \cite{Scribble}
which is based on multiparty session type (MPST) theory \cite{MPST}.
The Scribble protocol is used as a basis for generating
TypeScript APIs targeted for the Node.js server-side runtime
and React.js browser-side framework.
By implementing the generated APIs in their full-stack applications,
developers can enjoy static guarantees with respect 
to communication safety: 
there will not be any communication mismatches,
and channel linearity will be respected.

\subparagraph{S2: \newtheory}
We present an extension of the canonical MPST
theory that supports \textit{routed communication}.

@@

By step \textbf{S1}, we address the limitation \textbf{L1}.
Our work provides the same communication safety guarantees
as the current state-of-the-art, but 
is compatible with modern web programming tools and idioms.
Among the languages that compile to JavaScript,
we select TypeScript as it is arguably the most intuitive
to use, given it is a \textit{superset} of JavaScript 
\cite{UnderstandingTypeScript}.
It provides developers
with type-safety through its gradual, structural type system,
and is also interoperable with existing JavaScript libraries
that the developer may incorporate into their application.
Whilst some \cite{MVU2020} point out this limits its usability for
encoding multiparty session types, we believe that the language
offers sufficient features that we can use to provide developers with
communication safety guarantees whilst preserving a flexible, natural
and idiomatic workflow.
The TypeScript APIs we generate also support idioms commonly
found in modern web development (such as callbacks and
asynchronous operations), and provides an error handling
mechanism tailored to the dynamic web-based environment
(where roles can disconnect prematurely).

In step \textbf{S2},
we show that routing communication interactions through an intermediate
role, such as a web server, preserves the communication
safety properties and the semantics from the original
interactions.
By extending \codegen to implement \newtheory,
our workflow can implement protocols of
arbitrary communication structures over a server-centric
network topology using WebSockets; 
this tackles limitation \textbf{L2}.

>>>>

The objective of this project is to provide
developers with a development workflow
that \textit{provides communication safety guarantees in modern
web programming} through multiparty session types.

We tackle the limitations of the current
state-of-the-art with two solutions:

@@

Developers specify their communication protocol using
the Scribble protocol description language \cite{Scribble}
which is based on multiparty session type (MPST) theory \cite{MPST}.
The Scribble protocol is used as a basis for generating
TypeScript APIs targeted for the Node.js server-side runtime
and React.js browser-side framework.
By implementing the generated APIs in their full-stack applications,
developers can enjoy static guarantees with respect 
to communication safety: 
there will \textit{not} be any \textbf{deadlocks} or
\textbf{communication mismatches},
and \textbf{channel linearity} will be respected.

\subparagraph{S2: \newtheory}
We present an extension of the canonical MPST
theory that supports \textit{routed communication}.
We prove that \newtheory also extends the communication safety
properties of the canonical MPST theory, so
processes that implement \newtheory also enjoy
guarantees of \textbf{deadlock-freedom}, \textbf{channel linearity},
and the absence
of \textbf{communication mismatches}.

@@

By solution \textbf{S1}, we address the limitation \textbf{L1}.
Our work provides the same communication safety guarantees
as the current state-of-the-art, but 
is compatible with modern web programming tools and idioms.
Among the languages that compile to JavaScript,
we select TypeScript as it is arguably the most intuitive
to use, given it is a \textit{superset} of JavaScript 
\cite{UnderstandingTypeScript}.
It provides developers
with type-safety through its gradual, structural type system,
and is also interoperable with existing JavaScript libraries
that the developer may incorporate into their application.
Whilst some \cite{MVU2020} point out this limits its usability for
encoding multiparty session types, we believe that the language
offers sufficient features that we can use to provide developers with
communication safety guarantees whilst preserving a flexible, natural
and idiomatic workflow.
The TypeScript APIs we generate also support idioms commonly
found in modern web development (such as callbacks and
asynchronous operations), and provides an error handling
mechanism tailored to the dynamic web-based environment
(where roles can disconnect prematurely).

In solution \textbf{S2},
we show that routing communication interactions through an intermediate
role, such as a web server, preserves the communication
safety properties and the semantics from the original
interactions.
By extending \codegen to implement \newtheory,
our workflow can implement protocols of
arbitrary communication structures over a server-centric
network topology using WebSockets; 
this tackles limitation \textbf{L2}.
